---
title: "Suicide Rates Analysis(1985 to 2016)"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
#install.packages('countrycode')
library(countrycode)

theme_set(theme_light()) #cause it'll look pretty 
```

## Suicide Rates Overview 1985 to 2016

**Data Source **
United Nations Development Program. (2018). Human development index (HDI). Retrieved from <http://hdr.undp.org/en/indicators/137506>

World Bank. (2018). World development indicators: GDP (current US$) by country:1985 to 2016. Retrieved from <http://databank.worldbank.org/data/source/world-development-indicators#>

[Szamil]. (2017). Suicide in the Twenty-First Century [dataset]. Retrieved from <https://www.kaggle.com/szamil/suicide-in-the-twenty-first-century/notebook>

World Health Organization. (2018). Suicide prevention. Retrieved from <http://www.who.int/mental_health/suicide-prevention/en/>

```{r importing data, echo= TRUE}
data <- read_csv("master.csv")
```

## Global Trend


```{r pressure, echo=TRUE}
#global trend
a<-data %>%
  group_by(year)%>%
  mutate(sum_year=sum(`suicides/100k pop`))

ggplot(a,aes(year,sum_year))+geom_line()+geom_point()+
  geom_line(col = "deepskyblue3", size = 1) + 
  geom_point(col = "deepskyblue3", size = 2) + 
  labs(title = "Global Suicides (per 100k)",
       subtitle = "Trend over time, 1985 - 2015.",
       x = "Year", 
       y = "Suicides per 100k") + 
  scale_x_continuous(breaks = seq(1985, 2015, 2)) + 
  scale_y_continuous(breaks = seq(10, 20))
```


**Insights**

1. Globally the Suicide rate has dropped compared to 1985


2. Suicide rate has gradually incereased from 1985 till 1995 where the global suicide rate was 
   all time high.


3. The Highest Drop at suicide rate was observed within 2015 and 2016 but this is observation might be incorrenct    due to lack of data points at that certain year.

## Cleaning the Data
Dropped the `suicides/100k pop` & `HDI for year` variable due to in efficient labeling.Furthermore some variables labling was renamed to make it more concise.
```{r cleaning }
tidy_data <-  data %>% 
  select(-c(`HDI for year`, `suicides/100k pop`)) %>%
  rename(gdp_for_year = `gdp_for_year ($)`, 
         gdp_per_capita = `gdp_per_capita ($)`, 
         country_year = `country-year`) %>%
  as.data.frame()

```



### Adding Continent

Added continent for each country in our observations via using the `countrycode` package.

```{r adding continent}
tidy_data$continent <- countrycode(sourcevar = tidy_data[, "country"],
                              origin = "country.name",
                              destination = "continent")

```


### Factoring the Colums 


```{r }
tidy_data$generation <- factor(tidy_data$generation, 
                          ordered = T, 
                          levels = c("G.I. Generation", 
                                     "Silent",
                                     "Boomers", 
                                     "Generation X", 
                                     "Millenials", 
                                     "Generation Z"))

tidy_data <- as_tibble(tidy_data)

```


## Suicides Per 100k in Each continent


```{r }
b<-tidy_data %>% group_by(continent) %>%
  summarize(population = sum(population), 
            suicides = sum(suicides_no), 
            suicides_per_100k = (suicides / population) * 100000)

ggplot(b, aes(continent,suicides_per_100k,fill=continent))+
         geom_bar(stat = "identity")
```




**Insights**
1. Europe has the highest rate of the Suicide.


2. Africa has the lowest rate of suicide


## Suicides per 100k in between each Generation


```{r each generation, fig.height=8 ,fig.width=8}

tidy_data %>% group_by(generation) %>%
  summarize(
    population = sum(population), 
    suicides = sum(suicides_no), 
    suicides_per_100k = (suicides / population) * 100000
  ) %>%
  ggplot(aes(generation,suicides_per_100k,fill=generation))+geom_bar(stat = "identity")
```


## Suicide rates in male-female


```{r }
tidy_data %>% group_by(sex) %>%
  summarize(
    population = sum(population), 
    suicides = sum(suicides_no), 
    suicides_per_100k = (suicides / population) * 100000
  ) %>%
  ggplot(aes(sex,suicides_per_100k,fill=sex))+geom_bar(stat = "identity")
```


## Trend over time, By Sex


```{r }
trend_sex<-tidy_data %>% 
  group_by(year,sex) %>%
  
  summarise(
    population = sum(population), 
    suicides = sum(suicides_no), 
    suicides_per_100k = (suicides / population) * 100000
  )

ggplot(trend_sex,aes(year,suicides_per_100k,color=sex),)+
  geom_line(size=1)+
  geom_point(size=2)+
  facet_wrap(~sex,nrow = 2,scales = 'free')+
  labs(title = "Trends Over Time, by Sex", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Sex") + 
  theme(legend.position = "none") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F)
```


## Global Suicides per 100k, by Country


```{r , fig.height=15,fig.width=8}
country <- tidy_data %>%
  group_by(country, continent) %>%
  summarize(n = n(), 
            suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  arrange(desc(suicide_per_100k))

country$country <- factor(country$country, 
                          ordered = T, 
                          levels = rev(country$country))

ggplot(country, aes(x = country, y = suicide_per_100k, fill = continent)) + 
  geom_bar(stat = "identity") + 
  
  labs(title = "Global suicides per 100k, by Country",
       x = "Country", 
       y = "Suicides per 100k", 
       fill = "Continent") +
  coord_flip() +
  scale_y_continuous(breaks = seq(0, 45, 2)) + 
  theme(legend.position = "bottom")
```


## GDP & Suicide Rate

We see from our previous analysis that Suicide Rate in europian countries are High.So is there any corelation between a country's GDP and suicide rate? 
Here each data point represents each country

```{r }
gdp_suicide_rate <- tidy_data %>%
  group_by(country,continent)%>%
  summarise(
    population = sum(population), 
    suicides = sum(suicides_no), 
    suicides_per_100k = (suicides / population) * 100000,
    gdp_per_capita = mean(gdp_per_capita))


gdp_suicide_rate<- gdp_suicide_rate %>% ungroup() %>%
  arrange(desc(suicides_per_100k)) 


ggplot(gdp_suicide_rate,aes(gdp_per_capita,suicides_per_100k,color=continent))+
  geom_point(aes(alpha=0.01))+
  geom_jitter()
```
It's Hard to see any trend with this scatter plot so we add a smooth line with `geom_smooth()`.

```{r smooth line, echo=FALSE}
gdp_suicide_rate_smooth <- tidy_data %>%
  group_by(country)%>%
  summarise(
    population = sum(population), 
    suicides = sum(suicides_no), 
    suicides_per_100k = (suicides / population) * 100000,
    gdp_per_capita = mean(gdp_per_capita))


gdp_suicide_rate_smooth<- gdp_suicide_rate %>% ungroup() %>%
  arrange(desc(suicides_per_100k)) 


ggplot(gdp_suicide_rate_smooth,aes(gdp_per_capita,suicides_per_100k))+
  geom_point(aes(alpha=0.01))+
  geom_jitter()+
  geom_smooth(method = "lm")

```
there's a slight positive corelation between the `gdp_per_capita` & `suicides_per_100k`

## Let's look into Asia


```{r}
gdp_suicide_rate_asia<-gdp_suicide_rate %>%
  filter(continent=="Asia")

ggplot(gdp_suicide_rate_asia,aes(gdp_per_capita,suicides_per_100k,color="red"))+
  geom_point(size=2)+
  geom_jitter() +
  geom_smooth(method = "lm")
```
Asia seems to have a negetive corelation between `gdp_per_capita` & `suicides_per_100k`





